Appendix B: Examples of Deep RL in Ecological Decision Problems

### 2021-06-04 appendices/appendix-B-examples.Rmd

@@

At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3]. 
Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
However, higher-level interfaces and high-level documentation are still relatively limited.
OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption. 
Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.

Although all the necessary tooling for RL is implemented in Python, the R language is more familiar to most ecologists.
Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.

In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
Alternately, users may prefer running the analysis inside a docker container.  
The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
This can be particularly useful for users running ML algorithms on remote servers, as configuring R, Python, and CUDA environments appropriately can be challenging.

Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.
Note that the RMarkdown source-code for this file can be found in the `appendices` directory of the project repository.
From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.

```{r message=FALSE}
#install.packages("renv")
renv::restore()
```

Once the packages have installed, we are ready to load the necessary libraries.  Note that the `import` function from `reticulate` package acts much like the `library` command in R, though it does not attach the package function to the global namespace.  To make it more convenient to access those functions, we can assign a shorthand name.

@@

## Python dependencies loaded via R
sb3         = import ("stable_baselines3")
gym         = import ("gym")
gym_fishing = import("gym_fishing")
gym_conservation = import("gym_conservation")

# 
source("../R/plotting-helpers.R")
```

@@

We begin by selecting an environment from `gym_fishing` by name, passing the optional parameter `sigma` to initialize the environment with multiplicative Gaussian environmental noise of sd of `0.1`.

```{r }
## initialize the environment
env = gym$make("fishing-v1", sigma = 0.1)
```

@@

```{r}
algos <- c("Optimal", "A2C", "TD3", "TD3_tuned")
sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
policy_df <- opt_policy  %>% mutate(model = factor("Optimal", algos))

@@

We will compare this optimal solution with results from the several different RL algorithms.
Unlike the optimal solution, the RL methods never know the underlying growth function used by the simulation (or even if such a function actually exists).
These methods merely seek to learn a strategy for setting harvest quotas which maximizes the cumulative reward the receive from the environment.
At the heart of each deep RL algorithm is a neural network or a collection of neural networks. The RL agent uses these networks to approximate a policy function and/or a value function. The policy function is the agent's mapping of states to actions -- i.e. the agent's strategy. Value functions attempt to evaluate the goodness of a state or state-action pair towards achieving the RL objective of maximizing cumulative rewards. 
Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyper-parameters_ while the parameters of a RL algorithm that are learned during training are called parameters. For example, the number of layers in a neural network is a hyper-parameter but a weight in the neural net is referred to as a parameter.
Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms,
for instance, the A2C algorithm originally described in @A2C.
Here, we train an agent using the A2C algorithm using default hyper-parameter settings with a multi-layer perceptron policy and value network composed of two 64-neuron layers.

```{r}
a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
```

@@

```{r}
if(!file.exists("../python/cache/a2c.zip")) {

  # Train the agent for a fixed number of timesteps
  a2c$learn(total_timesteps=300000L)

  # Save our trained agent for future use
  a2c$save("../python/cache/a2c")
}
```

Note that while default hyper-parameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyper-parameter selection. This is discussed further below. 
Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.  Note that a copy of the trained agents are included in the corresponding GitHub repository.

```{r}
# Simulate management under the trained agent
a2c = sb3$A2C$load("../python/cache/a2c")
```

@@

```{r}

## represent the initial state size in the 'rescaled' state space.
state = env$get_state( 0.75 )
state
```

@@

```{r}
a2c_sims = env$simulate(a2c, reps = 100L)
```

@@

```{r}
a2c_policy = env$policyfn(a2c, reps = 5L)
```

```{r}
sims_df <- a2c_sims %>% mutate(model = factor("A2C", algos)) %>% bind_rows(sims_df)
policy_df <- a2c_policy  %>% mutate(model = factor("A2C", algos))  %>% bind_rows(policy_df)

@@

## TD3

We train a second agent based on the TD3 algorithm.
Recall that the Twin Delayed Deep Deterministic Policy Gradient, commonly known as TD3 [@TD3], is a generalization of Deep Deterministic Policy Gradient, or DDPG, which extends the DQN algorithm @DQN to a continuous action space using the deterministic policy gradient. 
DDPG and DQN are also available in the stable baselines framework [@sb3].
One of the most salient differences here is that A2C uses an Actor-Critic strategy that combines policy gradient and value iteration approaches, while TD3 is a policy gradient approach.

We will start with the default hyperparameters, which correspond to those used in @TD3, where the algorithm was first introduced.

```{r}
td3_untuned = sb3$TD3('MlpPolicy', env, seed = seed)
```

Once again, after our agent is defined we are ready to train it:

```{r }
if(!file.exists("../python/cache/td3_untuned.zip")){

  td3_untuned$learn(total_timesteps=300000L)
  td3_untuned$save("../python/cache/td3_untuned")

With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:

```{r}
# Simulate management under the trained agent
td3_untuned = sb3$TD3$load("../python/cache/td3_untuned")
td3_sims_untuned = env$simulate(td3_untuned, reps = 100L)
td3_policy_untuned = env$policyfn(td3_untuned, reps = 5L)
```

```{r}
sims_df <- td3_sims_untuned %>% mutate(model = factor("TD3", algos)) %>% bind_rows(sims_df)
policy_df <- td3_policy_untuned  %>% mutate(model = factor("TD3", algos))  %>% bind_rows(policy_df)

plot_sims(sims_df) / ( plot_policy(policy_df) + plot_reward(sims_df))
```
The untuned TD3 policy tends to overfish relative to the optimal solution, but is less variable than the A2C solution.
While it is possible that with much more training, this TD3 algorithm could further improve, subsequent learning may be difficult.  
Note that the reward under TD3 is nevertheless often within the margin of error of the optimal policy, which may limit future learning since further policy exploration is more likely to reduce rewards.
Instead, we may be able to improve performance of TD3 by choosing different hyperparameters, which may allow it to imagine more possible policies or learn more effectively.

## Tuned TD3

Some of these hyper-parameters of the TD3 algorithm, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm. 
Others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance. 
For simplicity, we show only training with specified hyper-parameters here, found by various tuning experimentation (see <https://github.com/boettiger-lab/conservation-agents>) and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].

```{r}
# train an agent (model) on one of the environments:

policy_kwargs = list(net_arch=c(400L, 300L)) # "big"
# non-episodic action noise:
noise_std = 0.6656948079225263
n_actions = env$action_space$shape[0]
action_noise = sb3$common$noise$NormalActionNoise(
        mean=np$zeros(n_actions),
        sigma= noise_std * np$ones(n_actions)
        )

td3 = sb3$TD3( 'MlpPolicy', 
               env,  
               verbose=0L, 
               seed = seed,
               policy_kwargs = policy_kwargs,
               learning_rate = 0.0001355522450968401,
               gamma= 0.995, 
               batch_size = 128L,
               buffer_size = 10000L,
               train_freq = 128L, 
               gradient_steps = 128L,
               action_noise = action_noise
              )
```

We train this new TD3 agent:

```{r }
if(!file.exists("../python/cache/td3_tuned.zip")){

  td3$learn(total_timesteps=300000L)
  td3$save("../python/cache/td3_tuned")

@@

```{r}
# Simulate management under the trained agent
td3 = sb3$TD3$load("../python/cache/td3_tuned")
td3_sims = env$simulate(td3, reps = 100L)
td3_policy = env$policyfn(td3, reps = 5L)
```

```{r}
sims_df <- td3_sims %>% mutate(model = factor("TD3_tuned", algos)) %>% bind_rows(sims_df)
policy_df <- td3_policy  %>% mutate(model = factor("TD3_tuned", algos))  %>% bind_rows(policy_df)

@@

The tuned TD3 algorithm, though trained for an identical 300,000 timesteps as the untined one, settles on a policy much closer to the optimal policy.
The tuned TD3 agent tends to underfish on average, but only slightly, and avoids the occassional sharp drops in stock size seen when A2C occassionally sets too high a havest quota.
The tuned solution is still not optimal, but it is very close.
Moreover, the RL algorithm has been able to learn this nearly-optimal policy with no prior knowledge about the shape or nature of the population growth function used in the environment.

@@

```{r}
# Simulate management under the trained agent
env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
td3 = sb3$TD3$load("cache/td3")

@@

td3_state = td3_action = numeric(Tmax)
td3_state[1] = x0
td3_action[1] = NA

@@

  # RL-recommended harvest action:
  out = td3$predict(state)
  action = out[[1]]

  # Record state and resulting action
  td3_state[i] = env$get_fish_population(state)
  td3_action[i] = env$get_quota(action)

@@

df = bind_rows(
  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))

>>>>

Although all the necessary tooling for deep RL is most extensively implemented in Python, the R language is more familiar to most ecologists.
Fortunately, modern bindings such as the `reticulate` package [@reticulate] make it straightforward to use these tools without ever leaving the R interface.
In this appendix, we detail this "pure R" approach, as well as a "pure Python" approach.

In the R based-approach, R functions take responsibility from the user for translating commands into Python code before it is executed, an approach commonly referred to as meta-programming.
This still requires a local Python installation, which can be installed directly from R using the command `install_miniconda()` from the `reticulate` package in R.
Alternately, users may prefer running the analysis inside a docker container.  
The Rocker Project [@rocker] provides pre-built docker containers which include the necessary R and Python environments, as well as CUDA libraries required to take advantage of GPU-based acceleration on suitable architectures.
This can be particularly useful for users running ML algorithms on remote servers.

By using a cache of pre-trained agents distributed with this appendix, it should be possible to efficiently reproduce the results shown here on any recent laptop with the versions of software indicated.
Reproducing the training of RL agents is more computationally intensive, and a more powerful multi-core CPU or CUDA-enabled GPU is recommended.
Note that hardware differences are also a source of instability in the reproduciblity of RL training (see main text and Appendix A).

@@

Rather than implement deep RL algorithms from scratch, we illustrate the use of one of the leading software frameworks for research and application of deep RL methods.
At the time of writing, several major frameworks exist which provide reference implementations of the leading deep RL algorithms (see table 1 and Appendix A).
While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3]. 
Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
However, higher-level interfaces and high-level documentation are still relatively limited.
OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption. 
Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.

# Getting started

Clone the repository <https://github.com/boettiger-lab/rl-intro>, e.g. using the New Project->From Version Control->Git menu in RStudio.

From the project directory, we can then install all the necessary dependencies using `renv`, which helps ensure a reproducible environment of fixed versions of R packages and Python modules.

```{r message=FALSE}
#install.packages("renv")
#renv::restore()
```

Python users can install the dependencies listed in the repository's `requirements.txt` and run the corresponding python scripts found in the `python/` sub-directory instead. 
Meanwhile, the RMarkdown source-code for this file can be found in the `appendices` directory of the project repository.
Once the packages have installed, we are ready to load the necessary libraries.  Note that the `import` function from `reticulate` package acts much like the `library` command in R, though it does not attach the package function to the global namespace.  To make it more convenient to access those functions, we can assign a shorthand name.

@@

## Python dependencies loaded via R
sb3      = import("stable_baselines3")
torch    = import("torch")
gym      = import("gym")
gym_fish = import("gym_fishing")
gym_cons = import("gym_conservation")

source("../R/plotting-helpers.R")
```

@@

We begin by selecting an environment from `gym_fishing`.  While we use default parameter values for this example, note that is possible to pass alternative parameterizations using additional optional arguments.

```{r }
## initialize the environment
env = gym$make("fishing-v1")
```

@@

```{r}
algos <- c("Optimal", "PPO", "PPO_tuned")
sims_df <- opt_sims %>% mutate(model = factor("Optimal", algos))
policy_df <- opt_policy  %>% mutate(model = factor("Optimal", algos))

@@

We will compare this optimal solution with results from the several different RL algorithms.
Unlike the optimal solution, the RL methods never know the underlying growth function used by the simulation (or even if such a function actually exists).
These methods merely seek to learn a strategy for setting harvest quotas which maximizes the cumulative reward the receive from the environment.
At the heart of each deep RL algorithm is a neural network or a collection of neural networks. The RL agent uses these networks to approximate a policy function and/or a value function. The policy function is the agent's mapping of states to actions -- i.e. the agent's strategy. Value functions attempt to evaluate the goodness of a state or state-action pair towards achieving the RL objective of maximizing cumulative rewards. 
Confusingly, the parameters of a RL algorithm that detail the overall learning process are called _hyper-parameters_ while the parameters of a RL algorithm that are learned during training are called parameters. For example, the number of layers in a neural network is a hyper-parameter but a weight in the neural net is referred to as a parameter.
Before training an agent, we must first specify the hyper-parameters; the algorithm's parameters are often randomly initialized.
Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyper-parameters based on the original papers that introduced the corresponding algorithms.
For example, here we will try the Proximal Policy Optimization (PPO), originally described in @PPO.
Here, we train an agent using the PPO algorithm using default hyper-parameter settings with a multi-layer perceptron ('MlpPolicy') policy network and value network composed of two 64-neuron layers.

```{r}
ppo = sb3$PPO('MlpPolicy', env, seed = seed) 
```

@@

```{r}
if(!file.exists("../python/cache/ppo.zip")) {

  # Train the agent for a fixed number of timesteps
  ppo$learn(total_timesteps=300000L)

  # Save our trained agent for future use
  ppo$save("../python/cache/ppo")

Note that while default hyper-parameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyper-parameter selection. This is discussed further below. 
Having saved our trained agent to disk (`cache/ppo.zip`), we can then re-load this agent for evaluation or to continue training.  Note that a copy of the trained agents are included in the corresponding GitHub repository.

```{r}
# Simulate management under the trained agent
ppo = sb3$PPO$load("../python/cache/ppo")
```

@@

```{r}
## represent the initial state size in the 'rescaled' state space.
state = env$get_state( 0.75 )
state
```

@@

```{r}
ppo_sims = env$simulate(ppo, reps = 100L)
```

@@

```{r}
ppo_policy = env$policyfn(ppo, reps = 5L)
```

```{r}
# stack result simulations with those above and plot:
sims_df <- ppo_sims %>% 
  mutate(model = factor("PPO", algos)) %>% 
  bind_rows(sims_df)
policy_df <- ppo_policy  %>% 
  mutate(model = factor("PPO", algos))  %>% 
  bind_rows(policy_df)

@@

Note that the un-tuned version of PPO has found a viable policy which slowly increases its total reward over the course of the simulation,
but nevertheless is substantially sub-optimal in its performance.
Under the management of this agent, stock sizes average much higher than $B_{MSY}$ level of the optimal solution.
The stock sizes also fluctuate as the agent constantly adjusts the harvest quota in response to the previous level.

## Improving performance by tuning:

Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which uses standard function optimization approaches to vary the choice of hyper-parameters to determine which hyper-parameters maximize the agent's performance. 
For simplicity, we show only training with specified hyper-parameters here, found by various tuning experimentation (see <https://github.com/boettiger-lab/conservation-agents>) or the standard reference examples presented in `stable-baselines3-zoo` [@zoo].

```{r}
env = sb3$common$env_util$make_vec_env("fishing-v1", n_envs=4L, seed=seed)

# Create a fishing environment
# Create an agent
ppo_tuned = sb3$PPO("MlpPolicy",
                    "env" = env, 
                    "seed" = seed,
                    "batch_size" = 64L,
                    "n_steps" = 32L,
                    "gamma" = 0.9999,
                    "learning_rate" = 0.00038,
                    "ent_coef" = 0.0,
                    "clip_range" = 0.3,
                    "n_epochs" = 10L,
                    "gae_lambda" = 0.92,
                    "max_grad_norm" = 0.7,
                    "vf_coef" = 0.486355,
                    "policy_kwargs" = list(
                      "net_arch" = c(64L, 64L),
                      "activation_fn" = torch$nn$ReLU
                     )
                   )

We train this new PPO-based agent as before:

```{r }
if(!file.exists("../python/cache/ppo_tuned.zip")){

  ppo_tuned$learn(total_timesteps=300000L)
  ppo_tuned$save("../python/cache/ppo_tuned")

@@

```{r}
# Simulate management under the trained agent
ppo_tuned = sb3$PPO$load("../python/cache/ppo_tuned")
ppo_tuned_sims = env$simulate(ppo_tuned, reps = 100L)
ppo_tuned_policy = env$policyfn(ppo_tuned, reps = 5L)
```

```{r}
# stack result simulations with those above and plot:
sims_df <- ppo_tuned_sims %>% 
  mutate(model = factor("PPO_tuned", algos)) %>% 
  bind_rows(sims_df)
policy_df <- ppo_tuned_policy  %>% 
  mutate(model = factor("PPO_tuned", algos))  %>% 
  bind_rows(policy_df)

@@

The tuned solution is still not optimal, but it is very close.
Moreover, the RL algorithm has been able to learn this nearly-optimal policy with no prior knowledge about the shape or nature of the population growth function used in the environment.

@@

```{r}
# Simulate management under the trained agent
env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
agent = sb3$PPO$load("../python/cache/ppo_tuned")

@@

state = action = numeric(Tmax)
state[1] = x0
action[1] = NA

@@

  # RL-recommended harvest action:
  out = agent$predict(state, determinstic=TRUE)
  action = out[[1]]

  # Record state and resulting action
  state[i] = env$get_fish_population(state)
  action[i] = env$get_quota(action)

@@

df = bind_rows(
  tibble(year = years, state = state, action = action, model = "(tuned) PPO"),
  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))
